---
title: "tRNA array blood cell-type meth"
author: "Richard J. Acton"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_caption: yes
    number_sections: yes
    toc: yes
    toc_float: yes
    df_print: paged
    keep_md: yes
  html_notebook:
    fig_caption: yes
    number_sections: yes
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
    fig_caption: yes
    df_print: kable
bibliography: "`r normalizePath(params$bib)`"
csl: "`r normalizePath(params$csl)`"
link-citations: yes
linkcolor: blue
#runtime: shiny
editor_options: 
  chunk_output_type: inline
params:
  bib: "../library.bib"
  csl: "../genomebiology.csl"
  getRaw: FALSE
---

# Introduction

Looking for blood cell-type specific DNAm patterns in tRNA genes.

Data from: [GSE35069](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE35069) [@Reinius2012]

> Six healthy male blood donors, age 38 ± 13.6 years, were included in the study. From each individual, global DNA methylation levels were analyzed in whole blood, peripheral blood mononuclear cells (PBMC) and granulocytes as well as for seven isolated cell populations (CD4+ T cells, CD8+ T cells, CD56+ NK cells, CD19+ B cells, CD14+ monocytes, neutrophils, and eosinophils), n=60 samples analyzed in total.

# Set-up

## Libs

```{r libs}
suppressPackageStartupMessages({
	library(magrittr)
	library(tidyverse)
	library(here)
	library(Biobase)
	library(GEOquery)
	library(ComplexHeatmap)
})
```

## Env

```{r setup}
if(!dir.exists("data")) {
	dir.create("data")
}
if(!dir.exists("graphics")) {
	dir.create("graphics")
}
```

# Data Read-in

```{r}
bltRNAs <- read_tsv(
	"../tRNA-GtRNAdb/tRNAs-in-hg19-blacklist-v2.txt",
	col_names = "tRNAname",
	col_types = "c"
) %>% pull()
```

```{r}
tRNAprobes <- 
read_tsv(
	"../tRNA-GtRNAdb/450k_coresponding_hg19tRNAs.bed",
	col_names = c(
		"chr","start","end","probe",
		read_tsv(
			"../tRNA-GtRNAdb/std_tRNA_header.txt",
			col_types = cols(.default = col_character())
		) %>% colnames()
	)
)

tRNAprobes %>% nrow()
```

```{r}
tRNAprobes <- tRNAprobes %>% filter(!tRNAname %in% bltRNAs)
tRNAprobes %>% nrow()
```

```{r}
tRNAgeBB <- read_delim(
	delim = "\t",
	 col_names = "tRNAname",
	 col_types = "c",
	 file = "../tRNA-GtRNAdb/BB_GWS_tRNA.txt"
) %>%
	pull()
```

```{r}
swsBB23 <- read_delim(
	delim = "\t",
	col_names = "tRNAname",
	col_types = "c",
	file = "../tRNA-GtRNAdb/swsBB23.tsv"
) %>%
	pull()
```

```{r}
swsBB23bl <- swsBB23[!swsBB23 %in% bltRNAs]
```

```{r, eval = params$getRaw}
ctmeth <- GEOquery::getGEO("GSE35069")[[1]]
```

# pre-processing

```{r, eval = params$getRaw}
metadata <- ctmeth %>% pData()
#metadata
```

```{r, eval = params$getRaw}
meth <- exprs(ctmeth)
tRNAmeth <- meth[rownames(meth) %in% tRNAprobes$probe,]
```

```{r, eval = params$getRaw}
rm(ctmeth)
gc()
```

```{r, eval = params$getRaw}
saveRDS(
	metadata,
	"data/metadata_bl.Rds"
)
saveRDS(
	tRNAmeth,
	file = "data/tRNAmeth_bl.Rds"
)
```

```{r}
metadata <- readRDS(
	"data/metadata_bl.Rds"
)
tRNAmeth<- readRDS(
	file = "data/tRNAmeth_bl.Rds"
)
```

# Summary Info

Data processing: `r metadata$data_processing[1]`

Cell categories

```{r}
metadata %>%
	distinct(source_name_ch1)
```
Three categories: `r paste0(unique(metadata$source_name_ch1)[1:3],collapse = ", ")` have other cell-types on this list within them.

Methylation data available for `r nrow(tRNAmeth)` out of `r nrow(tRNAprobes)` probes, covering `r tRNAprobes %>% filter(probe %in% rownames(tRNAmeth)) %>% distinct(tRNAname) %>% nrow()` tRNAs. (Including `r tRNAprobes %>% filter(probe %in% rownames(tRNAmeth)) %>% distinct(tRNAname) %>% filter(tRNAname %in% tRNAgeBB) %>% nrow()` of the top `r length(tRNAgeBB)` from the MeDIP-seq).

```{r}
myeloidLymphoid <- data.frame(
	source_name_ch1 = unique(metadata$source_name_ch1),
	myeloidLymphoid = c(NA,"lymphoid","myeloid","lymphoid","lymphoid","myeloid","lymphoid","lymphoid","myeloid","myeloid")
)

metadata <- metadata %>%
	left_join(
		myeloidLymphoid,
		by = "source_name_ch1"
	)
```

## tRNA methylation density

```{r}
tRNAmeth %>% 
	as.data.frame() %>%
	rownames_to_column(var = "probe") %>%
	gather(-probe,key = "probe",value = "meth") %>%
	ggplot(aes(meth)) + 
	geom_density(aes(colour = probe),show.legend = FALSE)
```

# Plotting functions

Ensembl methylation colours

```{r}
col_fun <- circlize::colorRamp2(c(0, 0.5, 1), c("#ffff00", "#19ff00", "#0033cc"))
```

# initial inspection

```{r,fig.width=16,fig.height=20}
ComplexHeatmap::Heatmap(
	tRNAmeth,
	bottom_annotation = HeatmapAnnotation(type = metadata$`tissue/cell type:ch1`)
)
```

# Median meth by tRNA - all cell categories

```{r}
meanMethbytRNA <- 
tRNAmeth %>% 
	as.data.frame() %>%
	rownames_to_column(var = "probe") %>%
	left_join(tRNAprobes %>% select(tRNAname,probe),by = "probe") %>%
	group_by(tRNAname) %>%
	#summarise_if(is.double,median) %>%
	summarise_if(is.double,mean) %>%
	set_colnames(c("name",metadata$title)) %>%
	as.data.frame() %>%
	set_rownames(.$name) %>%
	select(-name)
	
meanMethbytRNAMat <- data.matrix(meanMethbytRNA) 
colnames(meanMethbytRNAMat) <- metadata$title

#meanMethbytRNAMat
```

```{r}
meanMethbytRNA %>% nrow()
meanMethbytRNA[!rownames(meanMethbytRNA) %in% bltRNAs,] %>% nrow()
```

```{r,fig.width=9,fig.height=12}
set.seed(42)
meanMethbytRNAMatHeatmap <- 
meanMethbytRNAMat %>%
ComplexHeatmap::Heatmap(
	.,
	bottom_annotation = HeatmapAnnotation(
		#"Cell-type" = metadata$`tissue/cell type:ch1`#,
		"Cell-type" = metadata$source_name_ch1,
		"Lineage" = metadata$myeloidLymphoid
		# annotation_legend_param = list(
		# #	"Cell-type" = list(
		# 		
		# 	)
		# #)
	),
	#annotation_legend_size()
	#annotation_legend_side = "bottom",
	#heatmap_legend_side = "bottom",
	#Legend(ncol = 1,title = "LEGEND"),
	
	column_title_gp = gpar(fontsize = 16, fontface = "bold"),
	column_title = "Mean methylation by tRNA\nSorted Blood Cell-types (GSE35069)",
	name = "Meth",
	#column_names_rot = 45,
	column_names_gp = gpar(fontsize = 5),
	row_names_gp = gpar(
		fontsize = 5,
		col = if_else(rownames(.) %in% swsBB23bl,"red","black")#tRNAgeBB
	),
	col = col_fun,
	row_title = "tRNA gene",
	na_col = "grey",
	#packLegend(heatmap_legend,"Cell-type",direction = "horizontal")
	#heatmap_legend_param = list(at = c(-2, 0, 2))
	
	# heatmap_width = unit(8.5,"inches"),
	# heatmap_height = unit(11.6,"inches")
)

meanMethbytRNAMatHeatmap
```

```{r}
png(
	filename = "./graphics/meanMethbytRNAMatHeatmap_bl.png",
	width = 9, height = 12, units = "in", res = 384
)
meanMethbytRNAMatHeatmap
dev.off()
```

# Median meth by tRNA - low level cell-categories

## filtering higher level groups

```{r}
metadataLow <- 
metadata %>%
	filter(!source_name_ch1 %in% c("Whole blood","PBMC","Granulocytes"))

tRNAmethLow <- tRNAmeth[,metadataLow$geo_accession]
```

```{r}
meanMethbytRNALow <- 
tRNAmethLow %>% 
	as.data.frame() %>%
	rownames_to_column(var = "probe") %>%
	left_join(tRNAprobes %>% select(tRNAname,probe),by = "probe") %>%
	group_by(tRNAname) %>%
	#summarise_if(is.double,median) %>%
	summarise_if(is.double,mean) %>%
	set_colnames(c("name",metadataLow$title)) %>%
	as.data.frame() %>%
	set_rownames(.$name) %>%
	select(-name)
	
meanMethbytRNAMatLow <- data.matrix(meanMethbytRNALow) 
colnames(meanMethbytRNAMatLow) <- metadataLow$title

#meanMethbytRNAMat
```

```{r,fig.width=9,fig.height=12}
set.seed(42)
meanMethbytRNAMatHeatmapLow <- 
meanMethbytRNAMatLow %>%
ComplexHeatmap::Heatmap(
	.,
	bottom_annotation = HeatmapAnnotation(
		#"Cell-type" = metadata$`tissue/cell type:ch1`#,
		"Cell-type" = metadataLow$source_name_ch1,
		"Lineage" = metadataLow$myeloidLymphoid
		# annotation_legend_param = list(
		# #	"Cell-type" = list(
		# 		
		# 	)
		# #)
	),
	#annotation_legend_size()
	#annotation_legend_side = "bottom",
	#heatmap_legend_side = "bottom",
	#Legend(ncol = 1,title = "LEGEND"),
	
	column_title_gp = gpar(fontsize = 16, fontface = "bold"),
	column_title = "Mean methylation by tRNA\nSorted Blood Cell-types (GSE35069)",
	name = "Meth",
	#column_names_rot = 45,
	column_names_gp = gpar(fontsize = 5),
	row_names_gp = gpar(
		fontsize = 5,
		col = if_else(rownames(.) %in% swsBB23bl,"red","black")#tRNAgeBB
	),
	col = col_fun,
	row_title = "tRNA gene",
	na_col = "grey",
	#packLegend(heatmap_legend,"Cell-type",direction = "horizontal")
	#heatmap_legend_param = list(at = c(-2, 0, 2))
	
	# heatmap_width = unit(8.5,"inches"),
	# heatmap_height = unit(11.6,"inches")
)

meanMethbytRNAMatHeatmapLow
```

```{r}
png(
	filename = "./graphics/meanMethbytRNAMatHeatmapLow_bl.png",
	width = 9, height = 12, units = "in", res = 384
)
meanMethbytRNAMatHeatmapLow
dev.off()
```

```{r,fig.width=7,fig.height=4}
set.seed(42)
meanMethbytRNAMatLowBB6 <- 
meanMethbytRNAMatLow[rownames(meanMethbytRNAMatLow) %in% swsBB23,] %>%
ComplexHeatmap::Heatmap(
	.,
	bottom_annotation = HeatmapAnnotation(
		"Cell-type" = metadataLow$source_name_ch1,
		"Lineage" = metadataLow$myeloidLymphoid
	),
	column_title_gp = gpar(fontsize = 12, fontface = "bold"),
	column_title = "Mean methylation by tRNA\nSorted Blood Cell-types (GSE35069)",
	name = "Meth",
	#column_names_rot = 45,
	column_names_gp = gpar(fontsize = 6),
	row_names_gp = gpar(
		fontsize = 8#,
		#col = if_else(rownames(.) %in% tRNAgeBB,"red","black")
	),
	col = col_fun,
	row_title = "tRNA gene",
	na_col = "grey",
)
meanMethbytRNAMatLowBB6
```

```{r}
png(
	filename = "./graphics/meanMethbytRNAMatLowBB6_bl.png",
	width = 7, height = 4, units = "in", res = 384
)
meanMethbytRNAMatLowBB6
dev.off()
```

```{r,fig.height=10,fig.width=5}
varBytRNA <- apply(meanMethbytRNAMatLow,1,var,na.rm = TRUE) %>%
	enframe(name = "tRNAname", value = "variance") %>%
	arrange(desc(variance)) %>%
	mutate(tRNAname = factor(tRNAname,levels = rev(tRNAname), ordered = TRUE)) %>%
	mutate(tRNAge = tRNAname %in% swsBB23)
varBytRNA

varBytRNA %>%
	head(30) %>%
ggplot(aes(tRNAname,variance)) + 
	geom_col(aes(fill = tRNAge)) + 
	coord_flip()
```

# Are the tRNAge genes significantly different in methylation between myeloid and lymphoid

```{r}
meanMethbytRNAMatLowLong <- 
meanMethbytRNAMatLow[,as.character(metadataLow$title)] %>%
	as.data.frame() %>%
	rownames_to_column(var = "tRNAname") %>%
	gather(-tRNAname,key = "title",value = "meth") %>%
	left_join(
		metadata %>% select(title,myeloidLymphoid),
		by = "title"
	)

meanMethbytRNAMatLowLong %>%
	group_by(myeloidLymphoid) %>%
	summarise(
		mean = mean(meth,na.rm = TRUE),
		median = median(meth,na.rm = TRUE)
	) %>%
	mutate(meanDiff = mean - lag(mean)) %>%
	mutate(medianDiff = median - lag(median))
```

```{r}
wt <- wilcox.test(
	meth ~ myeloidLymphoid,
	data = meanMethbytRNAMatLowLong,
	alternative = "greater",
	correct = FALSE,
	conf.int = TRUE
)
#summary(wt)
wt
```

```{R}
sprintf("%.2e", wt$p.value)

wt$conf.int

wt$estimate*100
```

```{r}
meanMethbytRNAMatLowLongBBAge <- 
meanMethbytRNAMatLowLong %>% filter(tRNAname %in% swsBB23) #tRNAgeBB

meanMethbytRNAMatLowLongBBAge %>%
	group_by(myeloidLymphoid) %>%
	summarise(
		mean = mean(meth,na.rm = TRUE),
		median = median(meth,na.rm = TRUE)
	) %>%
	mutate(meanDiff = mean - lag(mean)) %>%
	mutate(medianDiff = median - lag(median))
```

```{r}
wtBBage <- wilcox.test(
	meth ~ myeloidLymphoid,
	data = meanMethbytRNAMatLowLongBBAge,
	alternative = "greater",
	correct = FALSE,
	conf.int = TRUE
)
#summary(wt)
wtBBage
```

```{R}
sprintf("%.2e", wtBBage$p.value)

wtBBage$conf.int

wtBBage$estimate
```

# Session Info

```{r sessionInfo}
sessionInfo()
```

# References

<div id = "refs"></div>
