---
title: "Pooled Samples for Targeted Bis-seq: EPIC array data"
author: "Richard J. Acton"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_caption: yes
    number_sections: yes
    toc: yes
    toc_float: yes
  html_notebook:
    fig_caption: yes
    number_sections: yes
    toc: yes
    toc_float: yes
bibliography: "`r normalizePath(params$bib)`"
params:
  bib: "../library.bib"
---

# Intro

# Set-up

## Libs

```{r}
library(meffil)
library(tidyverse)
library(data.table)
library(EpiDISH)
```

```{r}
if(!dir.exists("out")) {
	dir.create("out", showWarnings = FALSE, recursive = TRUE)
}
if(!dir.exists("graphics")) {
	dir.create("graphics", showWarnings = FALSE, recursive = TRUE)
}
```

# Data Read-in

## Sample meta data

```{r}
sampleMetaData <- read_delim(
	file = "data/sampleMetaDataProcessed.tab", ##!! not public?
	delim = "\t",
	col_types = "cdcccdidc",
	col_names = TRUE
)
```

```{r}
# get the mean age of the samples in the pools
meanAgeByPool <- sampleMetaData %>% 
	group_by(pool) %>%
	summarise(age = mean(age)) %>%
	rename(Sample_Name=pool) # rename for combining with samplesheet
```

## Array Data

```{r}
samplesheet <- meffil.read.samplesheet(
	base = "data/", # dir with IDATs ## !! to be released on GEO
	pattern = "GC-CB-7769_Infinium_HD_EPIC_SampleSheet_310718_sex.csv"
)
```

add mean samples ages to samplesheet

```{r}
samplesheet <- left_join(samplesheet, meanAgeByPool, by = "Sample_Name")
```

# Intial QC

```{r}
qc.objects <- meffil.qc(
	samplesheet, cell.type.reference = "blood gse35069", verbose = TRUE
)
saveRDS(
	qc.objects,file = "data/data_gitIg/targetedBSseqEPICQCobjects.Rds"
)
```

```{r, eval = FALSE}
qc.objects <- readRDS(
	file = "data/data_gitIg/targetedBSseqEPICQCobjects.Rds"
)
```

```{r}
#meffil.qc.parameters()#see which qc parameters are available
qc.parameters <- meffil.qc.parameters(#set QC parameters
	beadnum.samples.threshold             = 0.1,
	detectionp.samples.threshold          = 0.1,
	detectionp.cpgs.threshold             = 0.1, 
	beadnum.cpgs.threshold                = 0.1,
	sex.outlier.sd                        = 5
)
qc.summary <- meffil.qc.summary(#summarise QC
	qc.objects,
	parameters = qc.parameters
)
```

## Generate QC report

```{r}
meffil.qc.report(
	qc.summary,
	output.file=paste0("out/qc-report_NEW.html")
)
```

```{r}
outlier <- qc.summary$bad.samples
table(outlier$issue)
index <- outlier$issue %in% c(
	"Control probe (dye.bias)", 
	"Methylated vs Unmethylated",
	"X-Y ratio outlier",
	"Low bead numbers",
	"Detection p-value",
	"Sex mismatch",
	"Control probe (bisulfite1)",
	"Control probe (bisulfite2)"
)

outlier <- outlier[index,]
length(qc.objects)
# qc.objects <- meffil.remove.samples(qc.objects, outlier$sample.name)
# length(qc.objects)
```

# Raw beta values

```{r}
raw.betas <- meffil.load.raw.data(qc.objects)
saveRDS(raw.betas,file = "data/targetedBSseqEPICrawBetas.Rds")
```

```{r, eval = FALSE}
raw.betas <- readRDS(file = "data/targetedBSseqEPICrawBetas.Rds")
```

```{r}
write.table(
	raw.betas, file = "data/array/raw_betas.csv", sep = ",",
	col.names = TRUE, row.names = TRUE, quote = FALSE
)
```

### Raw beta density by pool

```{r}
rawBetaDensityByPool <- 
raw.betas %>% 
	as.data.frame() %>%
	rownames_to_column(var = "probe") %>%
	gather(key = "pool", value = "beta", `Pool 1`:`Pool 8`) %>%
	ggplot(aes(beta)) + 
		geom_density(aes(colour = pool))

ggsave(rawBetaDensityByPool,filename = "graphics/rawBetaDensityByPool.png")
rawBetaDensityByPool
```

# Cell Counts

## Houseman

```{r getCellCounts}
getCellCounts <- function(qc.objects) {
	cc <- t(sapply(qc.objects, function(obj) obj$cell.counts$counts))
	cc <- data.frame(Sample_Name = row.names(cc), cc, stringsAsFactors = FALSE)
	return(cc)
}

cc <- getCellCounts(qc.objects)
colnames(cc) <- c(colnames(cc)[1],paste0("Houseman_", colnames(cc)[2:7]))
samplesheet <- left_join(samplesheet, cc, by = "Sample_Name")
```

## EpiDISH

```{r}
#library(EpiDISH)
data(centDHSbloodDMC.m)
bloodRef1 <- centDHSbloodDMC.m[,1:6] #B-cells, CD4+ T-cells, CD8+ T-cells, NK-cells, Monocytes and Granulocytes (not Neutrophils and Eosinophils components of Granulocytes to minimise covars)
epiDISHres <- epidish(raw.betas, bloodRef1, method = "RPC") 
#epiDISHres$estF 
colnames(epiDISHres$estF) <- paste0("EpiDISH_", colnames(epiDISHres$estF))
EpiDISHcellTypeCounts <- epiDISHres$estF %>% 
	as.data.frame() %>% rownames_to_column(var="Sample_Name")

samplesheet <- left_join(samplesheet, EpiDISHcellTypeCounts, by = "Sample_Name")
```

```{r}
#normalizePath("../data/samplesheetCellCounts.tab")
write_tsv(samplesheet, path = "data/samplesheetCellCounts2.tab")
```

## DNAm age prediction

```{r}
DNAmAgeRes <- read.csv(
	"data/clockProbesBetas.output.csv"
)
DNAmAgeRes$ChronAge <- meanAgeByPool$age
DNAmAgeRes
```

```{r, fig.width = 6, fig.height = 6}
max <- max(DNAmAgeRes$DNAmAge, DNAmAgeRes$ChronAge)
r2 <- cor(DNAmAgeRes$DNAmAge, DNAmAgeRes$ChronAge)**2
DNAmAgeVsChronAge <- DNAmAgeRes %>% 
	select(ChronAge, DNAmAge, SampleID) %>%
	ggplot(aes(ChronAge, DNAmAge)) + 
		geom_abline(slope = 1, colour = "red") +
		geom_point(size=4, aes(shape = SampleID)) + 
		scale_shape_manual(values = rep(c(3,4),4)) +
		ylim(0, max)+
		xlim(0, max)+
		geom_smooth(method = "lm") + 
		guides(shape = FALSE) +
		labs(
			title = "Mean age of pool Vs predicted Age",
			subtitle = paste0(
				"R² = ", sprintf("%.4g", r2), " | Red line: y = 1x + 0"
			),
			y = "DNAmAge / yrs",
			x = "Chronological Age / yrs"
		)
ggsave(
	DNAmAgeVsChronAge,
	filename = "graphics/DNAmAgeVsChronAge.png",
	width = 6, height = 6
)
DNAmAgeVsChronAge
```

# Normalise betas

## Number of PCs of the technical variation matrix to use in functional normalisation

```{r}
pcPlot <- meffil.plot.pc.fit(qc.objects)
pcPlot$plot
```

```{r}
pc <- 2
```

```{r}
names(qc.objects[[1]]$quantiles)
norm.objects <- meffil.normalize.quantiles(qc.objects, number.pcs = pc)
```

## Normalisation

```{r}
norm.beta <- meffil.normalize.samples(
	norm.objects, cpglist.remove = qc.summary$bad.cpgs$name
)
saveRDS(norm.beta, file = "data/targetedBSseqEPICnormBetas.Rds")
```

```{r, eval = FALSE}
norm.beta <- readRDS(file = "data/targetedBSseqEPICnormBetas.Rds")
```

### normalised beta density by pool

```{r}
normBetaDensityByPool <- 
norm.beta %>% 
	as.data.frame() %>%
	rownames_to_column(var = "probe") %>%
	gather(key = "pool",value = "beta",`Pool 1`:`Pool 8`) %>%
	ggplot(aes(beta)) + 
		geom_density(aes(colour = pool))
ggsave(normBetaDensityByPool, filename = "graphics/normBetaDensityByPool.png")
normBetaDensityByPool
```

# tRNA probes

```{r}
# tRNAprobeIntersection <- read.table(
# 	file = "../tRNA-GtRNAdb/EPIC_probesBytRNA_full.txt",
# 	sep = "\t", header = TRUE
# )
```

## probes in tRNA +/- 100bp

```{r}
tRNAprobeIntersection <- read_tsv(
	file = "../tRNA-GtRNAdb/EPIC_coresponding_hg19tRNAs-100bp53.bed",
	col_names = c(
		"pChr","pStart","pEnd","probe","tChr",
		"tStart","tEnd","tRNAname","score",
		"strand","thickStart","thickEnd","RGB",
		"blockCount","blockSizes","blockStarts"
	)
)

```

```{r}
GWStRNAs <- unlist(
	read.table(
		"../tRNA-GtRNAdb/GenWideSigWintRNAs.txt",
		sep = "\t", stringsAsFactors=FALSE
	),
	use.names = FALSE
)
# add a column to the data with 'tRNAge' gene set membership
tRNAprobeIntersection$GWS <- tRNAprobeIntersection$tRNAname %in% GWStRNAs
```

```{r}
# get a characer vector of probes in tRNA genes
tRNAprobes <- unlist(tRNAprobeIntersection$probe, use.names = FALSE)
```

```{r}
# number of probes on EPIC overlapping tRNAs
tRNAprobes %>% length()
# number of tRNA genes with EPIC probes
tRNAprobeIntersection$tRNAname %>% unique() %>% length()
```

# tRNA betas

```{r}
# combine betas and tRNA data for probes in tRNA genes
tRNAprobesWithBetas <- left_join(
	by = "probe",
	tRNAprobeIntersection %>% mutate(probe = as.character(probe)),
	norm.beta[rownames(norm.beta) %in% tRNAprobes,] %>% as.data.frame() %>% 
		rownames_to_column(var = "probe")
) %>% 
	gather(key = "pool", value = "beta", `Pool 1`:`Pool 8`) %>% # betas in long format
	select(-score, -RGB, -blockCount, -blockSizes, -blockStarts) %>% 
	extract(
		tRNAname, c("nmt", "aa", "codon"),
		"(\\w*)-?tRNA-(i?\\w{3})(?:\\w+)?-(\\w+)-", remove = FALSE
	) %>%
	na.omit()
```

```{r}
# add age data to tRNA and betas
tRNAprobesWithBetas <- left_join(
	tRNAprobesWithBetas, meanAgeByPool %>% rename(pool = Sample_Name), by="pool"
)
```

## tRNAge loci methylation

```{r, fig.width = 16, fig.height = 9}
GWStRNAbetasPlot <- tRNAprobesWithBetas %>%
	filter(GWS == TRUE) %>%
	ggplot(aes(age, beta, group = tRNAname)) +
		#geom_line(aes(colour=aa))
		geom_point(aes(colour = probe)) +
		geom_smooth(method = "lm", aes(colour = tRNAname), se = TRUE) + 
		facet_wrap(~ tRNAname + tChr, scales = "free") + 
		guides(colour = FALSE)
ggsave(
	GWStRNAbetasPlot, filename = "graphics/GWStRNAbetasPlot.png",
	width = 16, height = 9
)
GWStRNAbetasPlot
```

# Session Info

```{r}
sessionInfo()
```

# References
